---
title: "Reproducible Research - Peer Assessment 1"
author: "Ron Ferens"
date: "Saturday, July 18, 2015"
output: html_document
---

## Data  
The variables included in this dataset are:

- **steps:** Number of steps taking in a 5-minute interval (missing values are coded as NA)

- **date:** The date on which the measurement was taken in YYYY-MM-DD format

- **interval:** Identifier for the 5-minute interval in which measurement was taken

## Loading and preprocessing the data

Loading the measured (raw) data:

```{r}
rawData <- read.csv("C:/Users/rferens/Git/datasciencecoursera/Reproducible.Research/Project.01/Data/activity.csv")
summary(rawData)
```

## What is mean total number of steps taken per day?

In order to make sure our initial processing is not affected by NA values, we will remove the incomplete entires and generate a sub-set with complete entires: (Notice there are no NA entires)

```{r}
data <- rawData[complete.cases(rawData),]
summary(data)
```

1. Calculate the total number of steps taken per day

```{r}
totalNumberOfStepsPerDay <- tapply( data$steps, data$date, sum)
```

2. Make a histogram of the total number of steps taken each day

```{r}
hist(totalNumberOfStepsPerDay,
     xlab="Total number of steps taken each day",
     main="Histogram of the total number of steps taken each day")
```

3. Calculate and report the mean and median of the total number of steps taken per day

```{r, results='asis'}
mean(totalNumberOfStepsPerDay, na.rm=TRUE)
median(totalNumberOfStepsPerDay, na.rm=TRUE)
```

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
averageNumberOfStepsTakenPerInterval <- tapply( data$steps, data$interval, mean)
plot(averageNumberOfStepsTakenPerInterval, type="l",
     main="Average number of steps taken",
     ylab="Averaged across all days",
     xlab="5-minute interval")
```